Rapid technological progress in white light-emitting diodes \\ and its sources in innovation and technology spillovers

### 2023-01-26 2_SSL_EES/article/article_main.tex
updated main document and supplementary information

\begin{figure*}[h]
 \centering
 \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/history_efficacy.pdf}
 \caption{Historical development of the luminous efficacy ($K$) of the most widely-used lighting technologies in human history from 1850 to 2020. Data shown for best performers by year of market introduction, with dashed lines. Dashed lines indicate the average improvement calculated from a 3rd-order polynomial fit to the data. The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_max^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al. \cite{Murphy2012}. The magnified plot shows the progress in cool white LEDs from 1996 to 2020, with the dashed line indicating a linear rate of efficacy improvement of 10lm/W per year. For comparison, efficacies of best performers in legacy lighting technologies for 2020 are shown as coloured horizontal lines. Note the logarithmic scale of the vertical axis on the main plot and the linear scale on the magnified plot. Abbreviations: HID - High-Intensity Discharge; CFL - Compact Fluorescent Lamp; Hal. - Halogen, Incd. - Incandescent. Source: own synthesis of published data based on a visual approach proposed by Azevedo et al. \cite{azevedo2009transition}. See Supplementary Information, section 4 (Complete List of Sources for Figures), for the full list of sources and references.}
 \label{fgr:history_efficacy}
\end{figure*}

@@

\begin{figure}
\centering
  \includegraphics[height=6.5cm]{2_SSL_EES/article/figures/cost_lamp_small.pdf}
  \caption{Historical development of retail sales prices (in 2020 USD) per luminous flux of LED-based luminaires, including light bulbs, spotlights and recessed lights from 2008 to 2020. Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S. Department of Energy (DOE) \cite{council2013assessment}. Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs. Source: own synthesis of data on LED sales prices collected from various consumer watchdog databases and publications. Data on CFL sales prices collected from \cite{eger2018origin} as well as various consumer watchdog databases and publications. Incandescent light bulb price is assumed constant based on the average in the covered time period. See Supplementary Information, section 4 (Complete List of Sources for Figures), for the full list of sources and references.}
  \label{fgr:cost_lamp_small}
\end{figure}

These  dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies  \cite{weinold2020long}\cite{Mills2014}\cite{Stegmaier2021}\cite{grubb2021new}. As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year for the EU \cite{eu2019impactass} and 442 TWh/year for the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced annually by all solar photovoltaic installations in these regions. Notably, market adoption of LED lighting is not limited to developed economies \cite{Kamat2020}. For example, durability, low up-front cost and high efficiency of LED light sources have led to their wide-spread adoption in rural West African communities without access to grid electricity \cite{Bensch2017}. LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{o2019optical}, \cite{Wyatt2020}, watches and smartphones \cite{Bai2017}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.

@@

\begin{figure*}
 \centering
 \includegraphics[height=17cm]{2_SSL_EES/article/figures/chip_architecture_overview.pdf}
 \caption{Historical evolution of light-emitting diode chip architectures. Rows from top to bottom: classical chip with lateral current spreading; Osram’s thin GaN flip-chip (vertical) architecture; flip-chip architecture; chip-scale package flip-chip architecture. Shown are side views of chips without packages, along a cutaway line best suited to the features of each architecture. The cutaway surface is indicated by a red arrow with an eye on the overlaid top view of each chip. Note that the dimensions are not to scale, and smaller features are greatly exaggerated for clarity. Years indicated correspond to the earliest identified patent priority date. Black frames around certain designs indicate chip designs not brought to large-scale production. Chip architecture abbreviations: TF - Thin-Film; FC - Flip-Chip; CSP - Chip-Scale Package; Material abbreviations: GaN - Gallium Nitride, ITO - Indium Tin Oxide, MQW - Multiple Quantum Well. Source: adapted and compiled from multiple patents and industry publications. See Supplementary Information, section 4 (Complete List of Sources for Figures), for the full list of sources and references.}
 \label{fgr:chip_architecture_overview}
\end{figure*}

@@

\begin{figure*}[h!]
 \centering
 \includegraphics[width=17.5cm]{2_SSL_EES/article/figures/subefficiencies_progress.pdf}
 \caption{Historical improvements in light-emitting diode technology. Panels A1-A2 show they key metrics used to compute forward voltage efficiency $\eta_{Vf}$ and droop efficiency $\eta_{droop}$. Panels B1-B2 show the sub-efficiencies internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$. \newline
 \textbf{Panel A1:} Device forward voltage at a test current of I=350mA. The physical limit for a blue light wavelength of 450nm without electric pumping is shown for reference. Data points for devices released in 2020 by various manufacturers are shown in an inset plot.
 \textbf{Panel A2:} Efficiency droop at the test current of I=350mA.
 \textbf{Panel B1:} Internal quantum efficiency, for different chip architectures, by type of measurement used. Note that the artifact in the DOE Average around 2013 is due to a change in definition for internal quantum efficiency laid out in \cite{doe_ssl_multiyear_2013}. Measurement methods: PL - Photoluminescence\cite{Shim_2018}, EL - Electroluminescence\cite{Getty_2009};  ABC Model\cite{Karpov_2014}, Ideality Factor\cite{Masui_2010}. “Not reported” denotes data points where the IQE measurement technique was not reported. Whiskers indicate reported ranges, where applicable.
 \textbf{Panel B2:} Light extraction efficiency for different chip architectures, by type of specific technology used to improve light-extraction efficiency. Abbreviations: TFFC - Thin-Film Flip-Chip; FC - Flip-Chip; CSP - Chip-Scale Package; ITO - Indium Tin Oxide; ZnO - Zinc Oxide; PSS - Patterned Sapphire Substrate; electr. - electrode; text. - textured.
 Source: own synthesis of data from the full list of sources provided in the Supplementary Information, Section 4. Data on average device performance adapted from U.S. Department of Energy (DOE) Reports\cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_rnd_2016}\cite{doe_ssl_rnd_2018} and an Osram company report\cite{beale_leds_2015}.}
 \label{fgr:subeff}
\end{figure*}

@@

\begin{figure*}[h!]
 \centering
 \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2003.pdf}
 \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2010.pdf}
 \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2016.pdf}
 \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2020.pdf}
 \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020 (top to bottom panels), normalized to 1 Watt of electric power input (yellow bar on the left). Sub-efficiencies corresponding to each loss channel are listed below each column and described in Section \ref{sec:device_performance_metrics}. Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts). Percentages for red, green and blue loss channels indicate losses of remaining red/green/blue light energy. The corresponding LED architectures from Figure \ref{fgr:chip_architecture_overview} are shown for reference: 2003 (top panel) - flip-chip with YGAG phosphor; 2010 (bottom panel) - flip-chip with 258 phosphor; 2016 (top panel) - flip-chip with PFS phosphor; 2020 (bottom panel) - flip-chip with SALON phosphor. Details on phosphors are provided in Table \ref{tab:phosphors} and SI section XX. Overall LED package efficiency is $\eta_L = 5.8\%$ in 2003 and $\eta_L = 12.7\%$ in 2010. Abbreviations: Scat. = Scattering. Source (Efficiency): own elaboration based on data in Figure \ref{fgr:subeff}, as detailed in Section \ref{sec:methods}. Source (spectral data): Adapted from 2003\cite{Mueller2002}, 2010\cite{MuellerMach2005}, 2016\cite{Murphy2015}, 2019\cite{Hoerder2019}.}
 \label{fgr:waterfall}
\end{figure*}

@@

@@

@@

Our study has analysed in a systematic and granular way the sources of dramatic cost reductions and efficiency improvements in white light-emitting diodes since their introduction to the market in 1996. We find that the total LED device efficiency increased from $5.8\%$ to $38.7\%$ between 2003 and 2020 has been predominantly driven by LED technology innovations that have affected all physical energy loss channels and corresponding device sub-efficiencies. We also find that among those innovations, at least nine were driven by technology originating in areas of science and technology outside LEDs and solid-state lighting—i.e., those nine innovations can be referred to as technology spillovers . These spillovers were responsible for $8.5\%$ of the total efficiency improvements and nearly $100\%$ of the improvements in consumer experience metrics.

Our manufacturing cost model shows that a $95.5\%$ decrease in the cost of producing white classic-chip LEDs (from $1.11\$$ to $0.05\$$ in 2020 USD) between 2003 and 2020. While efficiency improvements have played an important role in these cost reductions, the largest components of such cost reductions have been driven by increases in the wafer size and yields across different manufacturing steps. In contrast with efficiency improvements, these increases were mainly a result of learning-by-doing and manufacturing equipment improvements rather than LED technology innovations or spillovers.

Our analysis of the sources, mechanisms and enablers of the identified technology spillovers which were significant drivers of efficiency improvements, highlights the critical role played by a growing deep understanding of the physical, chemical and optical phenomena underlying the operation of LEDs, as well as materials science and technology and nanotechnology involved in the production of LEDs. Specifically, deep physical understanding of LED device efficiency loss channels enabled important breakthroughs and spillovers in LEDs and will enable further innovations to improve LED and SSL efficiency in the future, as expected by eminent experts in the field \cite{Weisbuch2020}. This suggests that additional research in these areas and a more deliberate search for spillovers may accelerate these expected future advances. Our findings on spillover enablers show that these efforts can be further supported or even accelerated by knowledge exchange events and long-term partnerships between academia and industry, dedicated mission-driven public R\&D funding, and freedom of search in academia. This reinforces arguments made against the dichotomy of basic research versus applied research \cite{narayanamurti2016cycles} \cite{narayanamurti2021genesis} and the calls for open, inclusive and flexible research cultures \cite{Stephan2021}.

There are various important avenues of future research that are opened up by our analysis. First, future work could expand the cost model by collecting and including data for a broader set of chip architectures and analyzing the impact of individual innovations and spillovers on costs as opposed to performance. Second, a deeper dive on the role of learning-by-doing is needed both in the cost and performance analysis to identify different types of learning by doing and how it came about. Third, building on the work on LED sub-efficiencies and physical limits, future efforts could focus on identifying priority areas for further efficiency improvements in LEDs and SSL in general. Finally, by comparing the drivers of innovation, technology spillovers, cost reductions and performance improvements at a granular level across different clean energy technologies, we can identify patterns or differences that would help us formulate recommendations for industry and policymakers aimed at accelerating further clean energy innovation for climate change mitigation.

@@

\section*{Acknowledgements}
This research is supported by the grant from the Alfred P. Sloan Foundation titled \textit{“What factors drive innovation in energy technologies? The role of technology spillovers and government investment”}. Michael Weinold further gratefully acknowledges financial support from the Swiss Study Foundation. We would like to thank Venkatesh Narayanamurti, Gabriel Chan, Anna Goldstein, Didier Sornette, and participants of the SPIE West 2021 Conference and C-EENRG Seminar Series of the University of Cambridge for many helpful discussions and feedback. We also express our deep gratitude to all interviewees for their willingness to participate in this study and their invaluable contributions. Support by German consumer organization Stiftung Warentest in providing access to their archive is gratefully acknowledged.

>>>>

\begin{figure*}[h]
 \centering
 \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/history_efficacy.pdf}
 \caption{Historical development of the luminous efficacy ($K$) of the most widely-used lighting technologies in human history from 1850 to 2020. Data shown for best performers by year of market introduction, with dashed lines. Dashed lines indicate the average improvement calculated from a 3rd-order polynomial fit to the data. The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_max^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al. \cite{Murphy2012}. The magnified plot shows the progress in cool white LEDs from 1996 to 2020, with the dashed line indicating a linear rate of efficacy improvement of 10lm/W per year. For comparison, efficacies of best performers in legacy lighting technologies for 2020 are shown as coloured horizontal lines. Note the logarithmic scale of the vertical axis on the main plot and the linear scale on the magnified plot. Abbreviations: HID - High-Intensity Discharge; CFL - Compact Fluorescent Lamp; Hal. - Halogen, Incd. - Incandescent. Source: own synthesis of published data based on a visual approach proposed by Azevedo et al. \cite{azevedo2009transition}. See Supplementary Information, Section 5 (Complete List of Sources for Figures), for the full list of sources and references.}
 \label{fgr:history_efficacy}
\end{figure*}

@@

\begin{figure}
\centering
  \includegraphics[height=6.5cm]{2_SSL_EES/article/figures/cost_lamp_small.pdf}
  \caption{Historical development of retail sales prices (in 2020 USD) per luminous flux of LED-based luminaires, including light bulbs, spotlights and recessed lights from 2008 to 2020. Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S. Department of Energy (DOE) \cite{council2013assessment}. Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs. Source: own synthesis of data on LED sales prices collected from various consumer watchdog databases and publications. Data on CFL sales prices collected from \cite{eger2018origin} as well as various consumer watchdog databases and publications. Incandescent light bulb price is assumed constant based on the average in the covered time period. See Supplementary Information, Section 5 (Complete List of Sources for Figures), for the full list of sources and references.}
  \label{fgr:cost_lamp_small}
\end{figure}

These  dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies  \cite{weinold2020long}\cite{Mills2014}\cite{Stegmaier2021}\cite{grubb2021new}. As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year for the EU \cite{eu2019impactass} and 442 TWh/year for the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced annually by all solar photovoltaic installations in these regions. Notably, market adoption of LED lighting is not limited to developed economies \cite{Kamat2020}. For example, durability, low up-front cost and high efficiency of LED light sources have led to their wide-spread adoption in rural West African communities without access to grid electricity \cite{Bensch2017}. LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{o2019optical}\cite{Wyatt2020}, watches and smartphones \cite{Bai2017}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.

@@

\begin{figure*}
 \centering
 \includegraphics[height=17cm]{2_SSL_EES/article/figures/chip_architecture_overview.pdf}
 \caption{Historical evolution of light-emitting diode chip architectures. Rows from top to bottom: classical chip with lateral current spreading; Osram’s thin GaN flip-chip (vertical) architecture; flip-chip architecture; chip-scale package flip-chip architecture. Shown are side views of chips without packages, along a cutaway line best suited to the features of each architecture. The cutaway surface is indicated by a red arrow with an eye on the overlaid top view of each chip. Note that the dimensions are not to scale, and smaller features are greatly exaggerated for clarity. Years indicated correspond to the earliest identified patent priority date. Black frames around certain designs indicate chip designs not brought to large-scale production. Chip architecture abbreviations: TF - Thin-Film; FC - Flip-Chip; CSP - Chip-Scale Package; Material abbreviations: GaN - Gallium Nitride, ITO - Indium Tin Oxide, MQW - Multiple Quantum Well. Source: adapted and compiled from multiple patents and industry publications. See Supplementary Information, Section 5 (Complete List of Sources for Figures), for the full list of sources and references.}
 \label{fgr:chip_architecture_overview}
\end{figure*}

@@

\begin{figure*}[h!]
 \centering
 \includegraphics[width=17.5cm]{2_SSL_EES/article/figures/subefficiencies_progress.pdf}
 \caption{Historical improvements in light-emitting diode technology. Panels A1-A2 show they key metrics used to compute forward voltage efficiency $\eta_{Vf}$ and droop efficiency $\eta_{droop}$. Panels B1-B2 show the sub-efficiencies internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$. \newline
 \textbf{Panel A1:} Device forward voltage at a test current of I=350mA. The physical limit for a blue light wavelength of 450nm without electric pumping is shown for reference. Data points for devices released in 2020 by various manufacturers are shown in an inset plot.
 \textbf{Panel A2:} Efficiency droop at the test current of I=350mA.
 \textbf{Panel B1:} Internal quantum efficiency, for different chip architectures, by type of measurement used. Note that the artifact in the DOE Average around 2013 is due to a change in definition for internal quantum efficiency laid out in \cite{doe_ssl_multiyear_2013}. Measurement methods: PL - Photoluminescence\cite{Shim_2018}, EL - Electroluminescence\cite{Getty_2009};  ABC Model\cite{Karpov_2014}, Ideality Factor\cite{Masui_2010}. “Not reported” denotes data points where the IQE measurement technique was not reported. Whiskers indicate reported ranges, where applicable.
 \textbf{Panel B2:} Light extraction efficiency for different chip architectures, by type of specific technology used to improve light-extraction efficiency. Abbreviations: TFFC - Thin-Film Flip-Chip; FC - Flip-Chip; CSP - Chip-Scale Package; ITO - Indium Tin Oxide; ZnO - Zinc Oxide; PSS - Patterned Sapphire Substrate; electr. - electrode; text. - textured.
 Source: own synthesis of data from the full list of sources provided in the Supplementary Information, Section 5. Data on average device performance adapted from U.S. Department of Energy (DOE) Reports\cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_rnd_2016}\cite{doe_ssl_rnd_2018} and an Osram company report\cite{beale_leds_2015}.}
 \label{fgr:subeff}
\end{figure*}

@@

\begin{figure*}[h!]
 \centering
 \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2003.pdf}
 \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2010.pdf}
 \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2016.pdf}
 \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2020.pdf}
 \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020 (top to bottom panels), normalized to 1 Watt of electric power input (yellow bar on the left). Sub-efficiencies corresponding to each loss channel are listed below each column and described in Section \ref{sec:device_performance_metrics}. Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts). Percentages for red, green and blue loss channels indicate losses of remaining red/green/blue light energy. The corresponding LED architectures from Figure \ref{fgr:chip_architecture_overview} are shown for reference: 2003 (top panel) - flip-chip with YGAG phosphor; 2010 (bottom panel) - flip-chip with 258 phosphor; 2016 (top panel) - flip-chip with PFS phosphor; 2020 (bottom panel) - flip-chip with SALON phosphor. Details on phosphors are provided in Table \ref{tab:phosphors} and SI section XX. Overall LED package efficiency is $\eta_L = 5.8\%$ in 2003 and $\eta_L = 12.7\%$ in 2010. Abbreviations: Scat. = Scattering. Source (Efficiency): own elaboration based on data in Figure \ref{fgr:subeff}, as detailed in Section \ref{sec:methods}. Source (spectral data): Adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
 \label{fgr:waterfall}
\end{figure*}

@@

Our findings are further supported by sensitivity analysis, presented in the Supplementary Information section SI4, where we find that sensitivity of the model to variation in its main parameters decreases over time with the increase of the number of DPW. In SI section 5 we also provide a comparison of our model with past cost calculations and projections published by US DOE on the basis of the \textit{LEDCOM} cost model \cite{ledcomv2} and industry data reported to the DOE as part of SSL round tables \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}.

@@

\section{Discussion}

\subsection{Performance Improvements}

Our findings show that the overall white LED efficiency has increased from $\eta_L=5.8\%$ in 2003 to $\eta_L=38.7\%$ in 2020. This efficiency increase was predominantly the result of a series of innovations in the LED architecture and materials as well as changes in the manufacturing process. Our comparison of improvements in device sub-efficiencies between 2003 and 2020 shows that the overall efficiency improvement was not dominated by any single sub-efficiency channel. Instead, there has been consistent progress across all device sub-efficiencies in the recorded period. Some sub-efficiencies, such as forward voltage efficiency, internal quantum efficiency and light extraction efficiency, are now within $\sim~10\%$ of their respective physical limits. A marked trend is the increasing relative contribution of losses in channels with physical limits on efficiency due to the underlying physics, such as light conversion efficiency of phosphors determined by the Stokes shift. Spectral efficiency also remains at $\sim17\%$ from its physical limit, showing potential for further improvements. Research on improving LED performance across efficiency loss channels continues \cite{cho2017white}\cite{Weisbuch2020}.

We also find that there has been a significant progress in consumer experience metrics, which was driven by innovations in the means of conversion of blue light generated by conventional GaN LEDs into white light required for general illumination. The first commercial white LEDs produced by Nichia in 1996 used a YAG phosphor that could generate only cool white light \cite{BANDO1998}. After a series of breakthrough innovations in phosphors represented in Figure 8 and described in SI section 3, LEDs today can be tuned for high colour rendering performance, high colour saturation and a range of desirable colour temperatures.

Overall, we find that LED technology innovations made crucial contributions to the progress in both LED performance and consumer experience metrics across the entire period covered by our study. Our interviews have also revealed an important role of incremental process improvements and learning-by-doing\cite{WRIGHT_1936}\cite{Arrow_1962} in the progress in LED efficiency. However, further research is needed to quantify the contribution of learning by doing to this progress.

\subsection{Cost Reductions}

Our LED manufacturing cost model shows that the cost of manufacturing low-to-mid power GaN-based white LED packages with classical chip architecture at a U.S. location using state-of-the-art equipment has decreased by 95.5\% from 1.11\$(2020) in 2003 to 0.05\$(2020) in 2020. In contrast with LED performance improvements, where progress was driven mostly by LED technology innovations, the dramatic decline in LED manufacturing cost happened mostly due to economies of scale stemming from an increase in the wafer size and higher yield per manufacturing step. Both were enabled by advances in manufacturing equipment performance and incremental process improvements from learning-by-doing.

Notably, our bottom-up cost model is constructed to provide process-step resolution across three different key chip architectures: classical chips, flip chips, and chip-scale package flip chips. However, in this study we were able to collect data and compare the outcomes only for the classical chip architecture. Collecting the full set of data needed to populate the model for the remaining two architectures would require access to proprietary information from industry. With this limitation, tracking manufacturing cost declines across three key LED chip architectures remains a topic for future work.

\subsection{Contribution of Technology Spillovers}

We find that 9 technology spillovers identified in our study, listed in Table \ref{tab:spillovers}, affected all dimensions of white LED performance. Three spillovers associated with the use of YAG/YGAG phosphors in LEDs played the key role in the first commercial white LED lighting products, essentially enabling the solid-state lighting market and industry of today. Spillovers also had a significant effect on physical device performance, cumulatively contributing to 8.5\% of the total LED lamp efficiency improvements between 2003 and 2020. Technology spillovers were particularly important for consumer experience performance of white LED lighting sources, with corresponding spillover-driven innovations responsible for $\sim$100\% of the improvements in consumer experience metrics.

Following the framework for the analysis of technology spillovers proposed by Stephan and colleagues \cite{Stephan2021}, we gathered information from the interviews, historical records, literature reviews and industry publications to analyze the sources, mechanisms, and enablers of the identified spillovers into the white LED technology listed in Table 4. Among the spillover sources, we find that all 9 spillovers had origins in basic science disciplines such as various branches of chemistry, materials science, optics and photonics, and solid-state physics. Five spillovers also utilized technical knowledge and expertise in cathode ray tubes, fluorescent lighting, optoelectronic devices, nanotechnology, and nature-inspired material design.

Among the spillover mechanisms, six spillovers (involved in all phosphors except PFS and SLA, plus ITO) were a result of application of external scientific and technical knowledge already available to researchers and inventors. Three remaining spillovers (involved in PFS and SLA phosphors, plus PSS) occurred as an outcome of targeted search for relevant external knowledge outside the LED domain. In addition, at least two spillovers (involved in 258 and PFS phosphors) occurred through direct R\&D collaboration.

Among important enabling factors for the identified spillovers, we highlight public mission-driven R\&D funding; industry-academia partnerships; firm experience in multiple industries; conferences that brought together researchers from academia and industry; cultural and language proximity; freedom of search in academia; and university alumni networks.

We find that, on average, it took 26 years from the initial scientific discovery or invention to the moment of its spillover into the LED domain, with this time varying from 5 to almost 70 years. In contrast, it took much less time – just 6 years on average, varying from just a few months to 19 years – to develop a commercial application for the spillover knowledge in the LED market.

@@

Our study has analysed in a systematic and granular way the sources of dramatic cost reductions and performance improvements in white light-emitting diodes since their introduction to the market in 1996. We find that the total LED device efficiency increase from 5.8\% to 38.7\% between 2003 and 2020, as well as improvements in consumer experience metrics, have been predominantly driven by LED technology innovations that affected all physical energy loss channels and corresponding device sub-efficiencies. We also find that among those innovations, at least nine were driven by knowledge spillovers originating in areas of science and technology beyond LEDs or solid-state lighting. These spillover-driven innovations were responsible for 8.5\% of the total efficiency improvements and nearly 100\% of the improvements in consumer experience metrics.

Our manufacturing cost model shows a 95.5\% decrease in the cost of producing white classic-chip LEDs (from 1.11\$(2020) to 0.05\$(2020)) between 2003 and 2020, driven mostly by increases in the wafer size and yields across different manufacturing steps over time. In contrast with efficiency improvements, these increases were mainly a result of learning-by-doing and manufacturing equipment improvements rather than LED technology innovations or spillovers.

Our analysis of the sources, mechanisms and enablers of the identified technology spillovers which were significant drivers of improvements in LED efficiency and consumer experience metrics, highlights the critical role played by a  deep understanding of the physical, chemical and optical phenomena underlying the operation of LEDs, as well as materials science and technology and nanotechnology involved in the production of LEDs, for past and future advances in LED and solid-state lighting technology. Specifically, deep physical understanding of LED device efficiency loss channels enabled important innovations in LEDs and will continue enabling future improvements in LED and SSL efficiency, as expected by eminent experts in the field \cite{Weisbuch2020}. This suggests that additional research in these areas and a more deliberate search for relevant external knowledge may accelerate expected future advances in LED technology. These knowledge spillovers can be enabled or even accelerated by knowledge exchange events and long-term partnerships between academia and industry, dedicated mission-driven public R\&D funding, and a freedom of search in academia. This further reinforces arguments made against the dichotomy of basic research versus applied research \cite{narayanamurti2016cycles}\cite{narayanamurti2021genesis} and the calls for open, inclusive and flexible research cultures \cite{Stephan2021}.

There are various important avenues of future research that are opened up by our analysis. First, future work could expand the cost model by collecting and including data for a broader set of chip architectures and analyzing the impact of individual innovations and spillovers on the costs as opposed to performance. Second, a deeper dive on the role of learning-by-doing is needed both in the cost and performance analysis. Third, building on the work on LED sub-efficiencies and physical limits, future efforts could focus on identifying priority areas for further efficiency improvements in LEDs and SSL in general. Finally, by comparing the drivers of innovation, technology spillovers, cost reductions and performance improvements at a granular level across different clean energy technologies, we can identify patterns or differences that would help us formulate recommendations for industry and policymakers aimed at accelerating further clean energy innovation for climate change mitigation.

@@

\section*{Acknowledgements}
This research was supported by the grant from the Alfred P. Sloan Foundation titled \textit{“What factors drive innovation in energy technologies? The role of technology spillovers and government investment”}. Michael Weinold further gratefully acknowledges support from the Swiss Study Foundation. We would like to thank Venkatesh Narayanamurti, Gabriel Chan, Anna Goldstein, Didier Sornette, and participants of the SPIE West 2021 Conference and C-EENRG Seminar Series of the University of Cambridge for many helpful discussions and feedback. We also express our deepest gratitude to all interviewees for their willingness to participate in this study and share their invaluable insights with us.

### 2023-01-26 2_SSL_EES/article/article_supplementary.tex
updated main document and supplementary information

\newpage
\section{Manufacturing Cost Model}

@@

\begin{figure}[h]
	\centering
    \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/costmodel_calibration.pdf}
	\caption{Modelled manufacturing cost for light-emitting diode packages split by manufacturing step category. The top graph shows the data from LEDCOM model and round table discussions published by the United States Department of Energy (DoE). The bottom graph shows the results of the cost model developed in this thesis. Hatched bars are projections given by the sources. Inset plot shows the number of die per wafer (DPW) calculated for different wafer sizes used in the model.}
	\label{fig:costmodel_calibration}
\end{figure}

>>>>

\newpage
\section{Manufacturing Cost Model}
\label{sec:costmodel}

@@

\begin{figure}[h]
	\centering
    \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/costmodel_calibration.pdf}
	\caption{Modelled manufacturing cost for light-emitting diode packages split by manufacturing step category. The top graph shows the data from LEDCOM model and round table discussions published by the United States Department of Energy (DoE). The bottom graph shows the results of the cost model developed in this thesis. Hatched bars are projections given by the sources. Inset plot shows the number of die per wafer (DPW) calculated for different wafer sizes used in the model. Sources for top panel: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}. Sources for bottom panel: own elaboration based on the cost model described in Section \ref{sec:costmodel}. Sources for Inset Plot: own elaborations based on numerical approximations for die-per-wafer estimations provided in \cite{de2005investigation}, informed by expert interviews.}
	\label{fig:costmodel_calibration}
\end{figure}
